---
title: "RNAseq PCA Batch Effects"
author: "Xengie Doan"
date: '`r Sys.Date()`'
output:
  html_document:
    code_folding: hide
    highlight: haddock
    latex_engine: xelatex
    theme: cerulean
    toc: yes
    toc_depth: 4
    toc_float: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy.opts=list(width.cutoff=50), tidy=TRUE, fig.width=12, fig.height=8,fig.path = "analysis/figs/", dev=c('png', 'pdf'))

pkgs = c("DESeq2", "tidyverse", "RColorBrewer", "pheatmap", "ggplot2", "plotly", "biomaRt", "org.Hs.eg.db", "AnnotationDbi", "clusterProfiler", "reshape2" ) #, "profvis") for profiling
lapply(pkgs, library, character.only = TRUE)
```

## All cNF & MPNST samples
### PCA by tumor type and individual 
```{r}
load("data/vsd.Rda")
p <- plotPCA(vsd, intgroup=c("individualID", "tumorType", "transplantationType", "sex"), returnData= TRUE)
percentVar <- round(100 * attr(p, "percentVar"))
ggplot(p, aes(PC1, PC2, color = individualID, shape= tumorType, label = apply(cbind(vsd$individualID, coalesce(vsd$transplantationType,ifelse(vsd$isCellLine == "true", "cellLine", NA))), 1, function(x) paste(x[!is.na(x)], collapse = ".")) )) + geom_point(size=3) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed() + ggrepel::geom_text_repel()
```


### Remove batch effects
use limma removeBatchEffect and visualize in a PCA plot

#### removing consortium
```{r}
mat <- assay(vsd)
#consortium is a big batch effect
mat <- limma::removeBatchEffect(mat, vsd$consortium)

assay(vsd) <- mat

p <- plotPCA(vsd, intgroup=c("individualID", "tumorType", "transplantationType", "isCellLine"), returnData= TRUE)
percentVar <- round(100 * attr(p, "percentVar"))
ggplot(p, aes(PC1, PC2, color = individualID, shape= tumorType,label = apply(cbind(vsd$individualID, coalesce(vsd$transplantationType,ifelse(vsd$isCellLine == "true", "cellLine", NA))), 1, function(x) paste(x[!is.na(x)], collapse = ".")) )) + geom_point(size=3) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed() + ggrepel::geom_text_repel()
```

#### removing consortium and is cell line
```{r}
mat <- assay(vsd)
#consortium is a big batch effect
mat <- limma::removeBatchEffect(mat, vsd$consortium, vsd$isCellLine)
assay(vsd) <- mat

p <- plotPCA(vsd, intgroup=c("individualID", "tumorType", "transplantationType", "isCellLine"), returnData= TRUE)
percentVar <- round(100 * attr(p, "percentVar"))
ggplot(p, aes(PC1, PC2, color = individualID, shape= tumorType, label = apply(cbind(vsd$individualID, coalesce(vsd$transplantationType,ifelse(vsd$isCellLine == "true", "cellLine", NA))), 1, function(x) paste(x[!is.na(x)], collapse = ".")) )) + geom_point(size=3) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed() + ggrepel::geom_text_repel()
```

#### removing consortium and xeno
```{r}

mat <- assay(vsd)
vsd$transplantationType[is.na(vsd$transplantationType)] <- "notXenograft"

#consortium is a big batch effect
mat <- limma::removeBatchEffect(mat, vsd$consortium, vsd$transplantationType)

assay(vsd) <- mat

vsd$transplantationType[which(vsd$transplantationType == "notXenograft")] <- NA

p <- plotPCA(vsd, intgroup=c("individualID", "tumorType", "transplantationType", "isCellLine"), returnData= TRUE)
percentVar <- round(100 * attr(p, "percentVar"))

ggplot(p, aes(PC1, PC2, color = individualID, shape= tumorType, label = apply(cbind(vsd$individualID, coalesce(vsd$transplantationType,ifelse(vsd$isCellLine == "true", "cellLine", NA))), 1, function(x) paste(x[!is.na(x)], collapse = ".")) )) + geom_point(size=3) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed() + ggrepel::geom_text_repel()
```



####removing cell line and xenograft
```{r}
mat <- assay(vsd)
vsd$transplantationType[is.na(vsd$transplantationType)] <- "no"

#consortium is a big batch effect
mat <- limma::removeBatchEffect(mat, vsd$transplantationType, vsd$isCellLine)
assay(vsd) <- mat

p <- plotPCA(vsd, intgroup=c("individualID", "tumorType", "transplantationType", "isCellLine"), returnData= TRUE)
percentVar <- round(100 * attr(p, "percentVar"))
ggplot(p, aes(PC1, PC2, color = individualID, shape= tumorType, label = apply(cbind(vsd$individualID, coalesce(vsd$transplantationType,ifelse(vsd$isCellLine == "true", "cellLine", NA))), 1, function(x) paste(x[!is.na(x)], collapse = ".")) )) + geom_point(size=3) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed() + ggrepel::geom_text_repel()
```

## Filtered cNF & MPNST 
remove 2-009 Xenograft and 2-009 Cell Line because they are contaminated with mouse and filtered for protein coding genes only. 

### PCA by sample and tumor type
```{r}
load("data/vsd_remove2009contaminants_protCodingOnly.Rda")
p <- plotPCA(vsd, intgroup=c("individualID", "tumorType", "transplantationType", "sex"), returnData= TRUE)
percentVar <- round(100 * attr(p, "percentVar"))
ggplot(p, aes(PC1, PC2, color = individualID, shape= tumorType, label = apply(cbind(vsd$individualID, coalesce(vsd$transplantationType,ifelse(vsd$isCellLine == "true", "cellLine", NA))), 1, function(x) paste(x[!is.na(x)], collapse = ".")) )) + geom_point(size=3) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed() + ggrepel::geom_text_repel()
```

### PCA trying to remove batch effects
#### removing consortium
```{r}
load("data/vsd_remove2009contaminants_protCodingOnly.Rda")
mat <- assay(vsd)
#consortium is a big batch effect
mat <- limma::removeBatchEffect(mat, vsd$consortium)

assay(vsd) <- mat

p <- plotPCA(vsd, intgroup=c("individualID", "tumorType", "transplantationType", "isCellLine"), returnData= TRUE)
percentVar <- round(100 * attr(p, "percentVar"))
ggplot(p, aes(PC1, PC2, color = individualID, shape= tumorType, label = apply(cbind(vsd$individualID, coalesce(vsd$transplantationType,ifelse(vsd$isCellLine == "true", "cellLine", NA))), 1, function(x) paste(x[!is.na(x)], collapse = ".")) )) + geom_point(size=3) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed() + ggrepel::geom_text_repel()
```

#### removing consortium and is cell line
```{r}
load("data/vsd_remove2009contaminants_protCodingOnly.Rda")
mat <- assay(vsd)
#consortium is a big batch effect
mat <- limma::removeBatchEffect(mat, vsd$consortium, vsd$isCellLine)
assay(vsd) <- mat

p <- plotPCA(vsd, intgroup=c("individualID", "tumorType", "transplantationType", "isCellLine"), returnData= TRUE)
percentVar <- round(100 * attr(p, "percentVar"))
ggplot(p, aes(PC1, PC2, color = individualID, shape= tumorType, label = apply(cbind(vsd$individualID, coalesce(vsd$transplantationType,ifelse(vsd$isCellLine == "true", "cellLine", NA))), 1, function(x) paste(x[!is.na(x)], collapse = ".")) )) + geom_point(size=3) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed() + ggrepel::geom_text_repel()
```

####removing cell line and xenograft
```{r}
load("data/vsd_remove2009contaminants_protCodingOnly.Rda")
vsd$xenoCellLine <- paste(vsd$transplantationType, vsd$isCellLine, sep = ".")

mat <- assay(vsd)
vsd$transplantationType[is.na(vsd$transplantationType)] <- "no"

#consortium is a big batch effect
mat <- limma::removeBatchEffect(mat, vsd$transplantationType, vsd$isCellLine)
assay(vsd) <- mat

p <- plotPCA(vsd, intgroup=c("individualID", "tumorType", "transplantationType", "isCellLine", "xenoCellLine"), returnData= TRUE)
percentVar <- round(100 * attr(p, "percentVar"))

#rename xeno no back to NA
vsd$transplantationType <- gsub("^no", NA, vsd$transplantationType)

ggplot(p, aes(PC1, PC2, color = individualID, shape= tumorType, label = apply(cbind(vsd$individualID, coalesce(vsd$transplantationType,ifelse(vsd$isCellLine == "true", "cellLine", NA))), 1, function(x) paste(x[!is.na(x)], collapse = ".")) )) + geom_point(size=3) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed() + ggrepel::geom_text_repel()
# ggplotly()
```

#### removing consortium, xenograft, and is cell line
```{r}
load("data/vsd_remove2009contaminants_protCodingOnly.Rda")
mat <- assay(vsd)

vsd$xenoCellLine <- paste(vsd$transplantationType, vsd$isCellLine, sep = ".")

#consortium is a big batch effect
mat <- limma::removeBatchEffect(mat, vsd$consortium, vsd$xenoCellLine)
assay(vsd) <- mat

p <- plotPCA(vsd, intgroup=c("individualID", "tumorType", "transplantationType", "isCellLine", "xenoCellLine"), returnData= TRUE)
percentVar <- round(100 * attr(p, "percentVar"))
ggplot(p, aes(PC1, PC2, color = individualID, shape= tumorType, label = apply(cbind(vsd$individualID, coalesce(vsd$transplantationType,ifelse(vsd$isCellLine == "true", "cellLine", NA))), 1, function(x) paste(x[!is.na(x)], collapse = ".")) )) + geom_point(size=3) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed() + ggrepel::geom_text_repel()
```
